 

### 2020-12-03 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd

Rising temperatures are expected to have profound effects on forests; however, it is not well understood how responses will vary across forest strata.
Here, we synthesize evidence as to how environmental conditions and foliar traits vary across vertical gradients, shaping leaf temperatures, metabolism, and ultimately whole-tree growth and mortality.
Strong vertical microclimatic gradients imply that canopy leaves are exposed to more solar radiation and higher evaporative demand than understory leaves.
However, foliar traits relevant to shaping leaf temperature ($T_{leaf}$) and metabolism also vary strongly across height or light gradients.
Resulting $T_{leaf}$'s generally track air temperature ($T_{air}$) and do not vary dramatically across the vertical gradient, except when hydraulic limitations impede evaporative cooling such that $T_{leaf}$'s of sun-exposed canopy leaves become significantly elevated above both $T_{air}$ and understory $T_{leaf}$.
Correspondingly, while leaf metabolism generally increases with height across the vertical gradient, differences in thermal sensitivity, optimal temperatures, and thermal damage thresholds are modest.
The implications for tree performance are mixed: whereas thermal buffering may contribute to the disproportionate stress of larger trees during drought, emerging tree-ring evidence suggests that understory trees experience relatively greater stress from high $T_{air}$.
Additional research will be important for improving our spotty understanding of the thermal sensitivity of metabolism and tree growth across vertical forest gradients.
In the meantime, our findings imply that while large canopy trees are the most vulnerable to warming when combined with drought, understory trees may be more vulnerable under more mesic conditions.

@@

**When vertical gradients in the biophysical environment and leaf traits combine to shape leaf energy budgets, $T_{leaf}$ often differs little from $T_{air}$ and between understory and canopy (Fig. **2?**).**
For instance... 	(Bolstad et al. 1999).
Similarly, during the dry season in a tropical moist forest in Panama... (Rey-Sánchez et al. 2016).
**However, canopy leaves can reach much higher maximum $T_{leaf}$, and higher $T_{leaf} - T_{air}$, than understory leaves.**
-	(Slot et al. 2019) and refs therein
-	(Fauset et al. 2018)
Sun leaves can also be  cooled relative to $T_{air}$ more than shade leaves (Rey-Sánchez et al. 2016).

>>>>

Rising temperatures are expected to have profound effects on forests; however, it is not well understood how responses will vary across forest strata.
Here, we synthesize evidence as to how environmental conditions and foliar traits vary across vertical gradients, shaping leaf temperatures, metabolism, and ultimately whole-tree growth and mortality.
Strong vertical microclimatic gradients imply that canopy leaves are exposed to more solar radiation and higher evaporative demand than understory leaves.
However, foliar traits relevant to shaping leaf temperature ($T_{leaf}$) and metabolism also vary strongly across height or light gradients.
Resulting $T_{leaf}$'s exceed air temperature ($T_{air}$) under conditions of high solar radiation, low wind, and low stomatal conductance.
Differences are most pronounced when hydraulic limitations impede evaporative cooling such that $T_{leaf}$'s of sun-exposed canopy leaves become dramatically elevated above both $T_{air}$ and understory $T_{leaf}$.
While leaf metabolism generally increases with height across the vertical gradient, differences in thermal sensitivity, optimal temperatures, and thermal damage thresholds are modest.
The implications for tree performance are mixed: whereas thermal buffering may contribute to the disproportionate stress of larger trees during drought, emerging tree-ring evidence suggests that understory trees experience relatively greater stress from high $T_{air}$.
Additional research will be important for improving our spotty understanding of the thermal sensitivity of metabolism and tree growth across vertical forest gradients.
In the meantime, our findings imply that while large canopy trees are the most vulnerable to warming when combined with drought, understory trees may be more vulnerable under more mesic conditions.

@@

**When vertical gradients in the biophysical environment and leaf traits combine to shape leaf energy budgets, under some conditions $T_{leaf}$ tracks  $T_{air}$ and therefore differs little between understory and canopy (Fig. **2?**).**
In a broadleaf deciduous forest, nighttime $T_{leaf}$'s were similar in the canopy and understory	(Bolstad et al. 1999).
Similarly, during the dry season in a tropical moist forest in Panama... (Rey-Sánchez et al. 2016).
**However, canopy leaves can reach much higher maximum $T_{leaf}$, and higher $T_{leaf} - T_{air}$, than understory leaves.**
-	(Slot et al. 2019) and refs therein
-	(Fauset et al. 2018)
Sun leaves can also be  cooled relative to $T_{air}$ more than shade leaves (Rey-Sánchez et al. 2016). There's some hint of this in Bolstad et al. 1999.
